Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-07-25 jso.qmd
smoothness metric
Updated smoothness metrics subsection, with definitions and more details.

<!--  
\begin{assumption}\label{assum:my_assumption}
This is an assumption.
\end{assumption}
Assumption \ref{assum:my_assumption} is convenient.

@@

\begin{thm}\label{thm:my_theorem}
This is a theorem.
\end{thm}
Theorem \ref{thm:my_theorem} is important.
-->

Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction
technique aimed at identifying informative linear projections of data.
The method involves optimising an objective function known as the PP
index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the
criteria for what constitutes interesting or informative projections.
Given high-dimensional data $X \in \mathbf{R}^{n\times p}$ and an index
function $f(\cdot)$, PP finds the orthonormal basis
$A \in \mathbf{R}^{p \times d}$ that maximises the index value of the
projection, $Y = XA$:

@@

If a PP index function is evaluated at some random points, as is done at
the initialization stage of JSO, then these random index values can be
interpreted as a random field, indexed by a space parameter, namely the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process.

The distribution of a Gaussian process is fully determined by its mean
and covariance function. The smoothness property comes into play in the
definition of the covariance function: if a PP index is very smooth,
then two close projection angles should produce close index values
(strong correlation); by contrast, if a PP index is not very smooth,
then two close projection angles might give very different index values
(fast decay of correlations with respect to distance between angles).

Popular covariance functions are parametric positive semi-definite
functions, some of which have a parameter to capture the smoothness of
the Gaussian field. In particular, consider the Matérn class of
covariance functions, defined by

$$
K(u):=\frac{(\sqrt{2\nu}u)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}(\sqrt{2\nu}u)\ ,
$$

where $\nu>0$ is the smoothness parameter and where $\mathcal{K}_\nu$ is
the modified Bessel function. The Matérn covariance function can be
expressed analytically when $\nu$ is a half-integer, the most popular
values in the literature being $\frac{1}{2}$, $\frac{3}{2}$ and
$\frac{5}{2}$. The parameter $\nu$, called *smoothness parameter*,
controls the decay of the covariance function. As such, it is an
appropriate measure of smoothness of a random field.

In this context, the parameter $\nu$ is suggested as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as `GpGp`
[@guinness2021gpgp] or `ExaGeoStatR` [@abdulah2023large], to fit the
hyperparameters of a GP covariance function on data. In this project,
the `GpGp` package is used.

After fitting the smoothness parameter $\nu>0$, its value can be
interpreted as follows: the higher $\nu$, the smoother the index
function.

@@

\begin{defn}\label{def:proj-dist}
Let $A \in \mathbf{R}^{p \times d}$ be the d-dimensional projection basis and
$A^*$ be the optimal basis that achieves the maximum index value for a given
PP problem defined by the shape to find, data dimension,
and the index function. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}

We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

\begin{defn}\label{def:squintability}
Let $g$ be the function that maps the projection distance, $d = d(A, A^*)$, 
to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$.
The squintability of $f(\cdot)$, denoted $\varsigma(f)$, is defined as 
$$\varsigma(f) = c \times \max_{d} \Delta g(d) \times \arg \max_{d}  \Delta g(d)$$ 
where $c$ is a constant scaling factor, $\max_A \Delta g(x)$ represents the 
maximum change in the index value along the curve $g$, and $\arg \max_{d} \Delta g(x)$ represents the projection distance at which this largest gradient is attained.

\end{defn}
Definition \ref{def:squintability} is ...

>>>>

```{=html}
<!--  
\begin{assumption}\label{assum:my_assumption}
This is an assumption.
\end{assumption}
Assumption \ref{assum:my_assumption} is convenient.

@@

\begin{thm}\label{thm:my_theorem}
This is a theorem.
\end{thm}
Theorem \ref{thm:my_theorem} is important.
-->
```
Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction
technique aimed at identifying informative linear projections of data.
The method involves optimising an objective function known as the PP
index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the
criteria for what constitutes interesting or informative projections.
Given high-dimensional data $X \in \mathbf{R}^{n\times p}$ and an index
function $f(\cdot)$, PP finds the orthonormal basis
$A \in \mathbf{R}^{p \times d}$ that maximises the index value of the
projection, $Y = XA$:

@@

An classical way to describe the smoothness of a function is to identify
how many continuous derivatives of the function exist. This can be
characterized by Sobolev spaces [@adams2003sobolev].

```{=tex}
\begin{defn}\label{def:sobolev_space}
The Sobolev space $W^{k,p}(\mathbb{R})$ for $1\leq p\leq \infty$ is the set of all functions $f$ in $L^p(\mathbb{R})$ for which all weak derivatives $f^{(\ell)}$ of order $\ell\leq k$ exist and have a finite $L^p$ norm.
\end{defn}
```
The Sobolev index $k$ in Definition \ref{def:sobolev_space} can be used
to characterize the smoothness of a function: if $f\in W^{k,p}$, then
the higher $k$, the smoother $f$. While this Sobolev index $k$ is a
useful measure of smoothness, it can be difficult to compute or even estimate
in practice.

To obtain a computable estimator of the smoothness of the index function
$f$, we propose an approach based on random fields. If a PP index
function $f$ is evaluated at some random bases, as is done at the
initialization stage of JSO, then these random index values can be
interpreted as a random field, indexed by a space parameter, namely the
random projection basis. This analogy suggests to use this random
training sample to fit a spatial model. We propose to use a Gaussian
process equipped with a Matérn covariance function, due to the
connections between this model and Sobolev spaces, see for example
@porcu2024matern.

The distribution of a Gaussian process is fully determined by its mean
and covariance function. The smoothness property comes into play in the
definition of the covariance function: if a PP index is very smooth,
then two close projection bases should produce close index values
(strong correlation); by contrast, if a PP index is not very smooth,
then two close projection bases might give very different index values
(fast decay of correlations with respect to distance between bases). Popular covariance functions are parametric positive semi-definite
functions.  In particular, the Matérn class of
covariance functions has a dedicated parameter to capture the smoothness of
the Gaussian field.

```{=tex}
\begin{defn}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{u}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{u}{\ell}\right)\ , u\geq0\label{eq:matern}
\end{equation}
where $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function \citep[10.25]{DLMF}. A multivariate extension $K(u)$, $u\in\mathbb{R}^d$ can be obtained by products of univariate covariance functions \eqref{eq:matern}.
\end{defn}
```
The Matérn covariance function can be expressed analytically when $\nu$
is a half-integer, the most popular values in the literature being
$\frac{1}{2}$, $\frac{3}{2}$ and $\frac{5}{2}$ [@rasmussen2006gaussian]. The parameter $\nu$,
called *smoothness parameter*, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field. For example, @karvonen2023asymptotic showed that if a
function $f$ has a Sobolev index of $k$, then the smoothness parameter
estimate $\nu$ in \eqref{eq:matern} cannot be asymptotically less than
$k$. See the survey @porcu2024matern for additional results on the connection between the Matérn model and Sobolev spaces. An interesting result is that the asymptotic case
$\nu\rightarrow\infty$ coincides with the Gaussian kernel:
$K_\infty(u)=\exp(-u^{2}/2)$.

In view of these results, the parameter $\nu$ is suggested as a measure
of the smoothness of the PP index function by fitting a Gaussian process
prior with Matérn covariance on a dataset generated by random
evaluations of the index function, as done at the initialization stage
of the jellyfish search optimization. There exist several R packages,
such as `GpGp` [@guinness2021gpgp] or `ExaGeoStatR` [@abdulah2023large],
to fit the hyperparameters of a GP covariance function on data, which is
usually done by maximum likelihood estimation. In this project, the
`GpGp` package is used.

```{=tex}
\begin{defn}
Let $\mathbf{A}=[A_1, \ldots, A_N] \in (\mathbf{R}^{p \times d})^N$ be d-dimensional projection bases, let $\mathbf{y}=[f(XA_1),\ldots,f(XA_N)]$ be the corresponding PP index values, and let $\mathbf{K}=[K_\theta(A_{i},A_{j})]_{1\leq i,j\leq N}\in\mathbb{R}^{N\times N}$ be the Matérn covariance matrix evaluated at the input bases, where the vector $\theta$ contains all the parameters of the multivariate Matérn covariance function $K$ (smoothness, outputscale, lengthscales). The log-likelihood of the parameters $\theta$ is defined by 
\begin{equation}
\mathcal{L}(\theta)=\log p(\mathbf{y}\left|\mathbf{A},\theta\right.)=-\frac{1}{2}\mathbf{y}^{\top}(\mathbf{K}+\sigma^{2}\mathbf{I})^{-1}\mathbf{y}-\frac{1}{2}\mathrm{\log}(\det(\mathbf{K}+\sigma^{2}\mathbf{I}))-\frac{N}{2}\log(2\pi)\,.\label{eq:gp_log_likelihood}
\end{equation}
where the nugget parameter $\sigma$ is the standard deviation of the intrinsic noise of the Gaussian process.
The optimal parameters (including smoothness) are obtained by maximum log-likelihood
\begin{equation}
\theta^* = \underset{\theta}{\max}\mathcal{L}(\theta)
\end{equation}
The resulting optimal smoothness parameter $\nu$ is chosen as our smoothness metric.
\end{defn}
```

The value of the optimal smoothness parameter $\nu>0$ can be
naturally interpreted as follows: the higher $\nu$, the smoother the
index function.

@@

```{=tex}
\begin{defn}\label{def:proj-dist}
Let $A \in \mathbf{R}^{p \times d}$ be the d-dimensional projection basis and
$A^*$ be the optimal basis that achieves the maximum index value for a given
PP problem defined by the shape to find, data dimension,
and the index function. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```
We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

```{=tex}
\begin{defn}\label{def:squintability}
Let $g$ be the function that maps the projection distance, $d = d(A, A^*)$, 
to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$.
The squintability of $f(\cdot)$, denoted $\varsigma(f)$, is defined as 
$$\varsigma(f) = c \times \max_{d} \Delta g(d) \times \arg \max_{d}  \Delta g(d)$$ 
where $c$ is a constant scaling factor, $\max_A \Delta g(x)$ represents the 
maximum change in the index value along the curve $g$, and $\arg \max_{d} \Delta g(x)$ represents the projection distance at which this largest gradient is attained.

\end{defn}
```
Definition \ref{def:squintability} is ...

### 2024-07-25 jso.tex
smoothness metric
Updated smoothness metrics subsection, with definitions and more details.

\begin{keyword}
    projection pursuit \sep jellyfish search optimiser
(JSO) \sep optimisation \sep grand tour \sep high-dimensional data \sep 
    exploratory data analysis
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, sharp corners, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, left=2mm, toptitle=1mm, rightrule=.15mm, arc=.35mm, titlerule=0mm, coltitle=black, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, bottomrule=.15mm, breakable, leftrule=.75mm, bottomtitle=1mm, opacitybacktitle=0.6, colback=white]

@@

If a PP index function is evaluated at some random points, as is done at
the initialization stage of JSO, then these random index values can be
interpreted as a random field, indexed by a space parameter, namely the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process.

The distribution of a Gaussian process is fully determined by its mean
and covariance function. The smoothness property comes into play in the
definition of the covariance function: if a PP index is very smooth,
then two close projection angles should produce close index values
(strong correlation); by contrast, if a PP index is not very smooth,
then two close projection angles might give very different index values
(fast decay of correlations with respect to distance between angles).

Popular covariance functions are parametric positive semi-definite
functions, some of which have a parameter to capture the smoothness of
the Gaussian field. In particular, consider the Matérn class of
covariance functions, defined by

\[
K(u):=\frac{(\sqrt{2\nu}u)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}(\sqrt{2\nu}u)\ ,
\]

where \(\nu>0\) is the smoothness parameter and where
\(\mathcal{K}_\nu\) is the modified Bessel function. The Matérn
covariance function can be expressed analytically when \(\nu\) is a
half-integer, the most popular values in the literature being
\(\frac{1}{2}\), \(\frac{3}{2}\) and \(\frac{5}{2}\). The parameter
\(\nu\), called \emph{smoothness parameter}, controls the decay of the
covariance function. As such, it is an appropriate measure of smoothness
of a random field.

In this context, the parameter \(\nu\) is suggested as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as
\texttt{GpGp} \citep{guinness2021gpgp} or \texttt{ExaGeoStatR}
\citep{abdulah2023large}, to fit the hyperparameters of a GP covariance
function on data. In this project, the \texttt{GpGp} package is used.

After fitting the smoothness parameter \(\nu>0\), its value can be
interpreted as follows: the higher \(\nu\), the smoother the index
function.

@@

Let \(A \in \mathbf{R}^{p \times d}\) be the d-D projection basis and
\(A^*\) be the optimal basis achieving the maximum index value for a
given projection pursuit problem defined by the shape to find, data
dimension, and the index function \(f(.)\). The projection distance
\(d(A, A^*)\), is defined as the Frobenius norm \(\lVert . \rVert _F\),
between the squared difference of these two bases:

\begin{equation}\protect\hypertarget{eq-proj-dist}{}{d(A, A^*) = \lVert M \rVert _F = \sum_{ij} M_{ij}^2, \text{ where } M = AA^\prime - A^*A^{*\prime}\ .}\label{eq-proj-dist}\end{equation}

>>>>

\begin{keyword}
    projection pursuit \sep jellyfish search optimiser
(JSO) \sep optimisation \sep grand tour \sep high-dimensional data \sep 
    exploratory data analysis
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, opacityback=0, coltitle=black, arc=.35mm, toprule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, breakable, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomtitle=1mm, bottomrule=.15mm, titlerule=0mm, rightrule=.15mm, toptitle=1mm, left=2mm]

@@

An classical way to describe the smoothness of a function is to identify
how many continuous derivatives of the function exist. This can be
characterized by Sobolev spaces \citep{adams2003sobolev}.

\begin{defn}\label{def:sobolev_space}
The Sobolev space $W^{k,p}(\mathbb{R})$ for $1\leq p\leq \infty$ is the set of all functions $f$ in $L^p(\mathbb{R})$ for which all weak derivatives $f^{(\ell)}$ of order $\ell\leq k$ exist and have a finite $L^p$ norm.
\end{defn}

The Sobolev index \(k\) in Definition \ref{def:sobolev_space} can be
used to characterize the smoothness of a function: if \(f\in W^{k,p}\),
then the higher \(k\), the smoother \(f\). While this Sobolev index
\(k\) is a useful measure of smoothness, it can be difficult to compute
or even estimate in practice.

To obtain a computable estimator of the smoothness of the index function
\(f\), we propose an approach based on random fields. If a PP index
function \(f\) is evaluated at some random bases, as is done at the
initialization stage of JSO, then these random index values can be
interpreted as a random field, indexed by a space parameter, namely the
random projection basis. This analogy suggests to use this random
training sample to fit a spatial model. We propose to use a Gaussian
process equipped with a Matérn covariance function, due to the
connections between this model and Sobolev spaces, see for example
\citet{porcu2024matern}.

The distribution of a Gaussian process is fully determined by its mean
and covariance function. The smoothness property comes into play in the
definition of the covariance function: if a PP index is very smooth,
then two close projection bases should produce close index values
(strong correlation); by contrast, if a PP index is not very smooth,
then two close projection bases might give very different index values
(fast decay of correlations with respect to distance between bases).
Popular covariance functions are parametric positive semi-definite
functions. In particular, the Matérn class of covariance functions has a
dedicated parameter to capture the smoothness of the Gaussian field.

\begin{defn}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{u}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{u}{\ell}\right)\ , u\geq0\label{eq:matern}
\end{equation}
where $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function \citep[10.25]{DLMF}. A multivariate extension $K(u)$, $u\in\mathbb{R}^d$ can be obtained by products of univariate covariance functions \eqref{eq:matern}.
\end{defn}

The Matérn covariance function can be expressed analytically when
\(\nu\) is a half-integer, the most popular values in the literature
being \(\frac{1}{2}\), \(\frac{3}{2}\) and \(\frac{5}{2}\)
\citep{rasmussen2006gaussian}. The parameter \(\nu\), called
\emph{smoothness parameter}, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field. For example, \citet{karvonen2023asymptotic} showed that if
a function \(f\) has a Sobolev index of \(k\), then the smoothness
parameter estimate \(\nu\) in \eqref{eq:matern} cannot be asymptotically
less than \(k\). See the survey \citet{porcu2024matern} for additional
results on the connection between the Matérn model and Sobolev spaces.
An interesting result is that the asymptotic case
\(\nu\rightarrow\infty\) coincides with the Gaussian kernel:
\(K_\infty(u)=\exp(-u^{2}/2)\).

In view of these results, the parameter \(\nu\) is suggested as a
measure of the smoothness of the PP index function by fitting a Gaussian
process prior with Matérn covariance on a dataset generated by random
evaluations of the index function, as done at the initialization stage
of the jellyfish search optimization. There exist several R packages,
such as \texttt{GpGp} \citep{guinness2021gpgp} or \texttt{ExaGeoStatR}
\citep{abdulah2023large}, to fit the hyperparameters of a GP covariance
function on data, which is usually done by maximum likelihood
estimation. In this project, the \texttt{GpGp} package is used.

\begin{defn}
Let $\mathbf{A}=[A_1, \ldots, A_N] \in (\mathbf{R}^{p \times d})^N$ be d-dimensional projection bases, let $\mathbf{y}=[f(XA_1),\ldots,f(XA_N)]$ be the corresponding PP index values, and let $\mathbf{K}=[K_\theta(A_{i},A_{j})]_{1\leq i,j\leq N}\in\mathbb{R}^{N\times N}$ be the Matérn covariance matrix evaluated at the input bases, where the vector $\theta$ contains all the parameters of the multivariate Matérn covariance function $K$ (smoothness, outputscale, lengthscales). The log-likelihood of the parameters $\theta$ is defined by 
\begin{equation}
\mathcal{L}(\theta)=\log p(\mathbf{y}\left|\mathbf{A},\theta\right.)=-\frac{1}{2}\mathbf{y}^{\top}(\mathbf{K}+\sigma^{2}\mathbf{I})^{-1}\mathbf{y}-\frac{1}{2}\mathrm{\log}(\det(\mathbf{K}+\sigma^{2}\mathbf{I}))-\frac{N}{2}\log(2\pi)\,.\label{eq:gp_log_likelihood}
\end{equation}
where the nugget parameter $\sigma$ is the standard deviation of the intrinsic noise of the Gaussian process.
The optimal parameters (including smoothness) are obtained by maximum log-likelihood
\begin{equation}
\theta^* = \underset{\theta}{\max}\mathcal{L}(\theta)
\end{equation}
The resulting optimal smoothness parameter $\nu$ is chosen as our smoothness metric.
\end{defn}

The value of the optimal smoothness parameter \(\nu>0\) can be naturally
interpreted as follows: the higher \(\nu\), the smoother the index
function.

@@
